# Fractions

```{r echo=FALSE}

require_cancel <- if (knitr::pandoc_to("html")) {"$$\\require{cancel}$$"} else {""}

```

Fractions can be written in two ways:

- as decimals fractions, for example $0.5$, $0.25$ and $0.\dot{3}$.
- as vulgar fractions, the following fractions have the same values as the examples above, $\frac{1}{2}$, $\frac{1}{4}$ and $\frac{1}{3}$. Vulgar fractions consist of two parts. The top, or **numerator**, and the bottom, the **denominator**.

Vulgar fractions are useful in algebra. The next section looks at some techniques for dealing with them.

`r require_cancel`

## Simplifying

Fractions can be *cancelled down* or simplified by dividing the numerator and denominator by the same thing. For example:

$$
\begin{aligned} \frac{18}{24} &= \frac{3 \times 6}{4 \times 6} \\
&= \frac{3 \times \cancel{6}}{4 \times \cancel{6}} \\
&= \frac{3}{4}
\end{aligned}
$$

The same can be done with algebraic fractions. 

$$
\begin{aligned} \frac{4xy}{6x} &= \frac{2y \times 2x}{3 \times 2x} \\
&= \frac{2y \times \cancel{2x}}{3 \times \cancel{2x}} \\
&= \frac{2y}{3}
\end{aligned}
$$

Sometimes you'll need to factorise expressions in the fraction in order to cancel it down.

$$
\begin{aligned} \frac{10x^2 + 5x}{4x+2} &= \frac{5x \times 2x + 5x \times 1}{2 \times 2x + 2 \times 1} \\
&= \frac{5x(2x+1)}{2(2x+1)} \\
&= \frac{5x\cancel{(2x+1)}}{2\cancel{(2x+1)}} \\
&= \frac{5x}{2}
\end{aligned}
$$

Here are some practice questions.

```{r, echo=FALSE}
knitr::include_url("https://numbas.mathcentre.ac.uk/question/127104/algebra-cancelling-fractions/embed/?token=17a9a4e4-a12e-4563-affd-e77faf02f68e", height="800px")
```

:::{.callout-important}
## Warning!

It is tempting to want to make cancellations like this:
$$
\begin{aligned} \frac{2x^2}{3x+7} &= \frac{2x\cancel{x}}{3\cancel{x}+1} \\
&= \frac{2x}{3+7} \\
&= \frac{2x}{10} \\
&= \frac{x}{5}
\end{aligned}
$$
However, please don't do it, as it's just plain wrong! Lets let $x=3$ and substitute it into the original $\frac{2x}{3x+7}$ and into incorrectly simplified version $\frac{x}{5}$. If the algebra is correct it should give the same answer.

We claim:
$$
\frac{2x^2}{3x+7} = \frac{x}{5}
$$
but if we substitute $x=2$ into both sides we get:
$$
\begin{aligned} \frac{2(3)^2}{3(3)+7} &= \frac{(3)}{5} \\
\frac{2 \times 9}{9+7} &= \frac{3}{5} \\
\frac{18}{16} &= \frac{3}{5} \\
\frac{9}{8} &= \frac{3}{5}
\end{aligned}
$$
Which is nonsense!
:::

## Multiplication and division

Multiplication and division of fractions is, thankfully, really easy!

### Multiplicaiton

For multiplication you simply multiply the numerators and denominators together. After the multiplication you may be able to cancel down the fraction. Just like this:

$$
\begin{aligned} \frac{2}{5} \times \frac{3}{4} &= \frac{2 \times 3}{5 \times 4} \\
&= \frac{6}{20} \\
&= \frac{3 \times 2}{10 \times 2} \\
&= \frac{3 \times \cancel{2}}{10 \times \cancel{2}} \\
&= \frac{3}{10}
\end{aligned}
$$
:::{.callout-tip}
## Pro-tip

It is possible to cancel before multiplying. Here is the same example revisited:

$$
\begin{aligned} \frac{2}{5} \times \frac{3}{4} &= \frac{2 \times 3}{5 \times 4} \\
&= \frac{2 \times 3}{5 \times 2 \times 2} \\
&= \frac{\cancel{2} \times 3}{5 \times 2 \times \cancel{2}} \\
&= \frac{3}{10}
\end{aligned}
$$

This can be super useful when dealing with large numbers or complex algebraic fractions.
:::

### Division

We can change a division into a multiplication by remembering **keep, change, flip**. We keep the first fraction as it is. Change the division, $\div$, symbol to a multiplication, $\times$, and flip the last fraction - swap the places of the numerator and denominator. This is called taking the reciprocal of the fraction. For example:

$$
\begin{aligned} \frac{3}{7} \div \frac{5}{2} &= \frac{3}{7} \times \frac{2}{5} \\
&= \frac{3 \times 2}{7 \times 5} \\
&= \frac{6}{35}
\end{aligned}
$$

## Addition and subtraction

Addition and subtraction is easy if the denominators are the same. We just add the numerators together and the demoninator stays the same.

$$
\begin{aligned} \frac{2}{5} + \frac{1}{5} &= \frac{2+1}{5} \\
&= \frac{3}{5}
\end{aligned}
$$

If the denominators are different we must make equivalent fractions with a common denominators first. It's like simplification, or cancelling down, in reverse.

$$
\begin{aligned} \frac{2}{3} + \frac{2}{9} &= \frac{2 \times 3}{3 \times 3} + \frac{2}{9} \\
&= \frac{6}{9} + \frac{2}{9} \\
&= \frac{6 + 2}{9} \\
&= \frac{8}{9}
\end{aligned}
$$
