# Differentiation

We often want to be able to find the gradient of a curved line. For that we need a new technique, called differentiation, that will give us a rule (a new function) to work out the gradient at any point on the curve.

## The tangent to a curve

The gradient at a point on a curve is the same as the gradient of the tangent at that point. A tangent to a curve is a straight line that just touches curve at that point. Below is a picture of the tangent to the curve when $x=5$. You can open up the graph and move the point around with the slider. 

```{r, echo=FALSE}
knitr::include_url("https://www.desmos.com/calculator/ahtukw0jwz?embed", height="500px")
```

Notice that the gradient will change depending on which value of $x$ you use. 

## The rules of differentiation

Luckily finding the rule to get the gradient of a curve is straight forward. The language we use for this process is like this. When function is differentiated a new function, the derivative, is found. The derivative enables you to find the gradient. There are lots of ways write this in mathematical notation. Here are the most common.

| original function | derivative |
|:--:|:--:|
| $y$ | $\frac{dy}{dx}$ |
| $f(x)$ | $f'(x)$ |

$\frac{dy}{dx}$ is pronounced 'dee $y$ by dee $x$', and $f'(x)$ is read as 'f dash of $x$'.

The rule for differentiating polynomials (functions made up of adding different powers of $x$)is:

:::{.callout-note}
* if $y=ax^n$ then $\frac{dy}{dx} = anx^{n-1}$, or,
* if $f(x)=ax^n$ then $f'(x) = anx^{n-1}$
**Times by the power, then take one off the power**
:::

Here are some examples:

If $y = 3x^4$ then $\frac{dy}{dx} = 3 \times 4 \times x^{4-1} = 12x^3$

Multiple terms added together are differentiated one by one then added together:

$$
\begin{aligned} y &= 6x^3 + 2x^2 + 4x + 5 \\
 &= 6x^3 + 2x^2 + 4x^1 + 5x^0 \\
\frac{dy}{dx} &= 3\times6x^{3-1} + 2\times 2x^{2-1} + 1\times 4x^{1-1} + 0\times5x^{0-1} \\
&= 18x^2 + 4x^1 + 4x^{0} + 0 \\
&= 18x^2 + 4x + 4
\end{aligned}
$$

In the above example we've used the following mathematical facts:

* $x=x^1$, $x$ on it's own is $x^1$
* $x^0=1$, you can always multiply by $x^0$ since it's $1$
* $0 \times a = 0$ anything times zero is zero

The take away from this is that constant terms, terms without $x$ in, disappear, and terms with just $x$ in loose the $x$.

Try these questions to get to grips with the rules of differentiation.

```{r, echo=FALSE}
knitr::include_url("https://numbas.mathcentre.ac.uk/question/62756/differentiation-using-a-table-of-derivatives-2/embed/?token=f1dc60aa-32b9-4906-98f2-4ddde19c9b79", height="800px")
```

## Finding gradient at a point

To find the gradient at a point. Differentiate the original function and then substitute the $x$ value of the point into the derivative. 

For example to find the gradient when $x=3$ for the function $y=x^2$. We would differentiate and then substitute in $x=3$.

$$
\begin{aligned} y &= x^2 \\
\frac{dy}{dx} &= 2x \\
&= 2(3) \\
&= 2 \times 3 \\
&= 6
\end{aligned}
$$
So the gradient at $x=3$ on the curve $y=x^2$ is $6$.

::: {.content-visible when-profile="econ"}

:::: {.callout-caution}
## Economics example

If a firm's total cost of producing $Q$ units is:

$$
TC = 3Q^2 + 10Q + 50
$$

economists often want to know how much it costs to increase output by one unit when production is already at some specific level. This extra cost is called the **marginal cost** at that level. It is found by differentiating total cost with respect to quantity and evaluating at the output level in question:

$$
MC = \frac{dTC}{dQ} = 6Q + 10
$$

For example, if the firm is currently producing $Q = 5$ units, the marginal cost is:

$$
MC = 6(5) + 10 = 40
$$

This means that when output is 5 units, the next unit adds approximately £40 to total cost. The derivative gives the rate of change at a particular point; economists call that rate the marginal cost when it measures the cost of one more unit at the current output level.
::::

:::: {.callout-caution}
## Economics example

Cost functions in economics often involve fractional powers. Suppose a firm's total cost is:

$$
TC = 5Q^{1.5} + 100
$$

Using the power rule, the marginal cost is:

$$
MC = \frac{dTC}{dQ} = 1.5 \times 5 \times Q^{0.5} = 7.5Q^{0.5}
$$

Notice that the exponent $1.5$ becomes $0.5$. Because the cost function contains a fractional power, the gradient is different at every value of $Q$. This means marginal cost is not a single number; it depends on where you are on the curve. You must always specify the output level when you quote a marginal cost.

When $Q = 4$:

$$
MC = 7.5 \times 4^{0.5} = 7.5 \times 2 = 15
$$

So at an output of 4 units, the 5th unit adds approximately £15 to total cost.

```{r, echo=FALSE, fig.height=7, fig.width=5, fig.align='center', fig.alt='A graph showing a marginal cost curve with a fractional power, MC equals 7.5 times the square root of Q. The point at Q equals 4 has a value of MC equals 15.'}
Q <- seq(0, 10, length.out = 200)
MC <- 7.5 * Q^0.5
plot(Q, MC, type = "n", xlim = c(0, 10), ylim = c(0, 30),
     xlab = "Output, Q", ylab = expression(MC == 7.5*Q^{0.5}),
     main = "Marginal cost with a fractional power",
     xaxp = c(0, 10, 5), yaxp = c(0, 30, 5),
     axes = FALSE)
abline(h = seq(0, 30, by = 6), v = seq(0, 10, by = 2), col = "grey90", lty = "dotted")
axis(1, at = seq(0, 10, by = 2))
axis(2, at = seq(0, 30, by = 6), las = 1)
box()
lines(Q, MC, lwd = 2)
points(4, 15, col = "blue", pch = 19, cex = 1.3)
```
::::

:::
