# Logarithms

Logarithms, or logs for short, are the same as powers just written in another way. 

## Reverse of indices

:::{.callout-note}
## Key point:

If $a^y =x$ then $y = \log_{a} x$.
:::

$a$ is called the base of the logarithm. When dealing with logs it's often useful to think of a numerical example to keep the idea straight in your head. 

$$
\begin{aligned} 10^3 &= 1000 \\
3 &= \log_{10} 1000
\end{aligned}
$$
This is the same fact written in index notation and as a logarithm.

## Rules of logarithms

Just as there are rules when dealing with indices, there are the corresponding rules when dealing with logarithms too.

:::{.callout-note}
## Key point:

* $\log_{a} x + \log_{a} y = \log_{a} xy$
* $\log_{a} x - \log_{a} y = \log_{a}{\frac{x}{y}}$
* $\log_{a} x^n = n\log_{a} x$
:::

We can use these rules to manipulate algebraic expressions. For example, let's write the following as a single logarithm:

$$
\begin{aligned}
3\log_{10} 2 + \log_{10} 5 - \log_{10} 4 &= \log_{10} 2^3 + \log_{10} 5 -  \log_{10} 4 \\
&= \log_{10} 8 + \log_{10} 5 -  \log_{10} 4 \\
&= \log_{10} (8 \times 5)  -  \log_{10} 4 \\
&= \log_{10} 40  -  \log_{10} 4 \\
&= \log_{10} (\frac{40}{4}) \\
&= \log_{10} (10) \\
&= 1
\end{aligned}
$$
This is how it was done:

* First we used the power rule $\log_{a} x^n = n\log_{a} x$, 
* then the addition rule $\log_{a} x + \log_{a} y = \log_{a} xy$, 
* and finally, the subtraction rule $\log_{a} x - \log_{a} y = \log_{a}{\frac{x}{y}}$. 
* Then notice $\log_{10} (10)= 1$ since $10^1=10$.

Have a go at these simplification questions.

```{r, echo=FALSE}
knitr::include_url("https://numbas.mathcentre.ac.uk/question/98284/logarithms-simplifying-expressions-2/embed/?token=236a0071-445b-43bf-a8a9-9611b4dcb5ae", height="340px")
```

## Solving equations with logarithms in

For example, let's solve $3\log_{10} x + \log_{10} 2 = \log_{10} 250$. First we'll apply the power rule  $\log_{a} x^n = n\log_{a} x$, then the addition rule $\log_{a} x + \log_{a} y = \log_{a} xy$: 

$$
\begin{aligned}
3\log_{10} x + \log_{10} 2 &= \log_{10} 250 \\
\log_{10} x^3 + \log_{10} 2 &= \log_{10} 250 \\
\log_{10} 2x^3 &= \log_{10} 250
\end{aligned}
$$

Now since the two sides are equal the values inside the logarithm must be equal. We can then go ahead and solve the resulting equation as normal.

$$
\begin{aligned}
\log_{10} 2x^3 &= \log_{10} 250 \\
2x^3 &= 250 \\
x^3 &= 125 \\
x &= \sqrt[3]{125} \\
 &= 5
\end{aligned}
$$

Have a go at the following questions:

```{r, echo=FALSE}
knitr::include_url("https://numbas.mathcentre.ac.uk/question/98356/logarithms-solving-logarithmic-equations-6-sum-of-logarithms/embed/?token=4dafb5d1-ca07-4bc5-a59f-7b18b7ff8ed8", height="340px")
```

## Some important bases

Some bases in logarithms come up more than others, because of that some bases have their own notation.

### The natural logarithm

A logarithm that has $e$ as it's base is known as the natural logarithm and has it's own symbol.

:::{.callout-note}
## Key point:

$$
\log_{e} x = \ln{x}
$$
:::

### Base 10

A logarithm that has $10$ as it's base has it's own symbol.

:::{.callout-note}
## Key point:

$$
\log_{10} x = \log{x}
$$
:::

You just don't bother writing the base.

## Differentiating $\ln{x}$

The rule for differentiating $\ln{x}$ is:

:::{.callout-note}
## Key point:

if $y = k\ln{ax}$ then $\frac{dy}{dx}= \frac{k}{x}$.
:::

Use that rule to try the following questions.

```{r, echo=FALSE}
knitr::include_url("https://numbas.mathcentre.ac.uk/question/62763/differentiation-using-a-table-of-derivatives-3/embed/?token=187473c8-cd94-4a87-b1cc-5f134572c64b", height="800px")
```
