# Critical points and ptimisation

A functions critical points are where it's gradient is zero. Differentiation can be used to find these points, and, these points can tell us the maximum on minimum values of a function. This is useful when optimising systems, for example:

* optimise profits by finding a maximum
* optimise a journey cost by finding minimum fuel use

If a situation can be defined with a function, differentiation can be used to help optomise it.

## Critical points - turning points on a curve

Maximum and minimum values of a function happen at turning points on a graph. At a turning point on a graph the gradient is zero. 

```{r, echo=FALSE}
knitr::include_url("https://www.desmos.com/calculator/yge6iy1mwg?embed")
```

Notice that the gradient will change depending on which value of $x$. 

## The rules of differentiation

Luckily finding the rule to get the gradient of a curve is straight forward. The language we use for this process is like this. When function is differentiated a new function, the derivative, is found. The derivative enables you to find the gradient. There are lots of ways write this in mathematical notation. Here are the most common.

| original function | derivative |
|:--:|:--:|
| $y$ | $\frac{dy}{dx}$ |
| $f(x)$ | $f'(x)$ |

$\frac{dy}{dx}$ is pronounced 'dee $y$ by dee $x$', and $f'(x)$ is read as 'f dash of $x$'.

The rule for differentiating polynomials (functions made up of adding different powers of $x$)is:

:::{.callout-note}
* if $y=ax^n$ then $\frac{dy}{dx} = anx^{n-1}$, or,
* if $f(x)=ax^n$ then $f'(x) = anx^{n-1}$
**Times by the power, then take one off the power**
:::

Here are some examples:

If $y = 3x^4$ then $\frac{dy}{dx} = 3 \times 4 \times x^{4-1} = 12x^3$

Multiple terms added together are differentiated one by one then added together:

$$
\begin{aligned} y &= 6x^3 + 2x^2 + 4x + 5 \\
\frac{dy}{dx} &= 6x^3 + 2x^2 + 4x^1 + 5x^0 \\
&= 3\times6x^{3-1} + 2\times x^{2-1} + 1\times 4x^{1-1} + 0\times5x^{0-1} \\
&= 18x^2 + 2x^1 + 4x^{0} + 0 \\
&= 18x^2 + 2x + 4
\end{aligned}
$$
In the above example we've used the following mathematical facts:

* $x=x^1$, $x$ on it's own is $x^1$
* $x^0=1$, you can always multiply by $x^0$ since it's $1$
* $0 \times a = 0$ anything times zero is zero

The take away from this is that constant terms, terms without $x$ in, disappear, and terms with just $x$ in loose the $x$.

Try these questions to get to grips with the rules of differentiation.

```{r, echo=FALSE}
knitr::include_url("https://numbas.mathcentre.ac.uk/question/62756/differentiation-using-a-table-of-derivatives-2/embed/?token=f1dc60aa-32b9-4906-98f2-4ddde19c9b79", height="340px")
```

## Finding gradient at a point

To find the gradient at a point. Differentiate the original function and then substitute the $x$ value of the point into the derivative. 

For example to find the gradient when $x=3$ for the function $y=x^2$. We would differentiate and then substitute in $x=3$.

$$
\begin{aligned} y &= x^2 \\
\frac{dy}{dx} &= 2x \\
&= 2(3) \\
&= 2 \times 3 \\
&= 6
\end{aligned}
$$
So the gradient at $x=3$ on the curve $y=x^2$ is $6$.
